Learning with Mental Models: the case of overestimation

### 2023-10-25 Paper/draft.tex
added last session

\title{Learning with Mental Models: the case of overestimation}
\author{Jimena Galindo}
\date{October 15, 2023}

\begin{document}
\maketitle
\begin{abstract}
I design a framework and a laboratory experiment that allows for the
comparison of multiple theories of misspecified learning. I focus on a
framework with endogenous information and a data-generating process
ruled by two fundamentals: an ego-relevant parameter and a state. Within
this framework I study three forces that can lead to misspecified
beliefs: initial misspecifications, learning traps and biased updating.
I find that biased updating is the main driver of misspecified beliefs
in the lab. In addition, I vary the degree of ego-relevance of the
parameter by introducing a stereotype treatment. The data is consistent
with biased updating in both cases but for potentially different
reasons: when learning about themselves, subjects attribute successes to
their own ability and failures to luck. Instead, in the stereotype
treatment, they compensate for initial negative biases by
over-attributing positive signals to the ability of others. This
translates into similar observed choices but different dynamics in
beliefs.
\end{abstract}

@@

When the setting features learning traps, even an agent who incorporates
all information correctly, may fall into learning traps as outlined by
\citet{Hestermann2021}. These traps are characterized by a combination
of an incorrect belief and an optimal action which produce information
that confirms the incorrect belief. Once an agent falls into a trap, the
belief will be stable and even with a correctly specified model of
learning, they will not be able to abandon their misspecified beliefs.
If, the agent is dogmatic about their initial belief,
\citet{Heidhues2018} show that they will inevitably fall into a trap and
thus will be able to rationalize and sustain their initial
misspecification
belief.\footnote{\citet{Gotte2022} study the case of agents with dogmatic initial beliefs in a 
laboratory experiment.}

\citet{Ba2023} moves away from dogmatism and endows the agents with a
mechanism through which they can abandon incorrect beliefs; this allows
them to avoid falling into learning traps. To do so agents perform
Bayesian hypothesis tests that evaluate which is the more likely
parameter out of two possibilities. By doing this, she characterizes the
set of situations in which, even agents who consider alternative
paradigms, may become trapped.\footnote{A 
similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion.}

@@

The agent chooses an action \(a\in A\) and observes a noisy outcome
\(h\). The outcome is a function of the agent's type, the state, and the
action. In particular \(h = h(\theta^*, \omega^*, a) + \varepsilon\)
with \(h(\cdot)\) increasing in both \(\theta^*\) and \(\omega^*\), and
such that conditional on a pair of parameters \((\theta, \omega)\),
there is a unique optimal action. additionally,
\(\varepsilon\sim N(0, \sigma)\) is noise in the output.

After observing the outcome, the agent updates his beliefs about
\(\theta\) and \(\omega\) using some algorithm and moves on to the next
period. He repeats this process infinitely many times. I make the
simplifying assumption that the agent is myopic and chooses the action
that maximizes the payoff in each period. This assumption simplifies the
analysis and plays a role in whether an agent who updates their beliefs
using Bayes rule would learn the truth or not, however, for the main
theory discussed in this section, the results hold even when relaxing
this assumption. The behavior of a forward-looking agent is further
discussed in the conclusion.

A key notion in this setting is that of a self-defeating
equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in 
\citet{Esponda2016} to this setting with only one agent}. A
\emph{self-defeating equilibrium} is a belief and action pair such that
the agent's belief about their type is misspecified and the outcome
generated by the action is consistent with the misspecified belief. This
means that the average outcome under the true type and the true state
equals the average output the agent expects under the misspecified
belief. The agent's belief is said to be \emph{stable} when this
happens.

Within this framework, I consider two nested theories of belief
updating: the first one is a dogmatic modeler from \citet{Heidhues2018},
The second one is a switcher, as in \citet{Ba2023}. The dogmatic modeler
can be seen as a switcher with an infinitely sticky initial belief so
the two theories are nested. However, outside of the limiting case, they
produce different predictions about the agent's behavior. I discuss each
of them separately in what follows.

@@

Take the action space to be the same as the state space \(\Omega\) and
\(\Omega =A=[\underline{a}, \infty)\).

A dogmatic agent does not update their beliefs about \(\theta\),
instead, he holds a degenerate belief that places probability on
\(\hat{\theta}\), which is potentially misspecified. In this case, no
matter how much information he gathers against being of type
\(\hat{\theta}\), he will not update his beliefs about it. Any
discrepancies between the observed outcomes and his believed type are
incorporated using the Bayes rule to update his beliefs about
\(\omega\). \citet{Heidhues2018} show that, under certain assumptions on
the per-period utility,\footnote{The assumptions are that 
$u$ is twice continuously differentiable with: (i)$u_{aa}<0$ and $u_a(\underline{a},  \theta, \omega)>0>u_a(\Bar{a}, \theta, \omega)$, 
(ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{a\theta}<0$ and $u_{a\omega}>0$. Where $\underline{a}$ is the
minimal action and $\Bar{a}$ is the maximal one. 
 The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} a dogmatic
modeler will inevitably fall into a self-defeating equilibrium. The
equilibrium will be such that the outcomes they observe reinforce their
belief on \(\omega\) in such a way that as \(t\to\infty\) the agent will
be sure that the state is some \(\omega^{'}\) consistent with their
believed type and the observed data. In other words, they will be in a
self-defeating equilibrium with a stable belief that places probability
one on the incorrect parameters \((\hat{\theta}, \omega_{\infty})\)

The mechanism by which the dogmatic agent falls into the self-defeating
equilibrium is the following: Suppose the agent holds the misspecified
belief that they are type \(\hat{\theta}>\theta^*\). For any prior over
\(\omega\), the agent will be disappointed by the outcome. He expected a
gain of \(h(\hat{\theta}, \mathbb{E}(\omega), a)\) but instead observes
\(h(\hat{\theta}, \omega^*, a)\). There are two possible sources for the
disappointment: the first is that the realized state is lower than the
expected state; the second source is that the agent is of type
\(\theta^*\) and therefore, for all possible states, his gain will be
lower than what he expected. Because the agent is dogmatic, he will not
update his beliefs about \(\theta\) and as a consequence will attribute
the disappointment to the state being lower than expected. He will
continue to update in this way until he converges to a belief about
\(\omega\) that is stable. Such a belief will explain the observed
utility perfectly and allow the agent to rationalize his dogmatic belief
about \(\theta\). Under the assumptions of \citet{Heidhues2018}, there
is a unique value of \(\omega\) at which the belief is stable, I will
refer to such value as \(\omega_\infty\). This mechanism is illustrated
in Example 1.

@@

Where \(h^t\) is the history of outcomes up to time \(t\) and \(\alpha\)
is the switching
threshold.\footnote{Notice that if $\alpha \to \infty$, the behavior of the switcher will be indistinguishable from that of the Dogmatic modeler. 
In this sense, the switcher is a generalization of the dogmatic type.}
By keeping track of the likelihood ratio, the agent can perform a
\emph{Bayesian hypothesis test} and adopt the Dogmatic belief that best
fits the
data.\footnote{In a related problem \citet{Schwarstein2021} proposes a similar updating procedure which relies on the Bayesian 
hypothesis test. However, in their model there is a sender who optimally chooses to propose a model that fits the data}

@@

A Bayesian agent simultaneously updates their beliefs about \(\theta\)
and \(\omega\) by using Bayes' rule. The posterior at period t+1 about
\(\theta\) after observing outcom \(o\) is given by: \[
p_{t+1}(\theta, \omega| h^t) = \frac{p[o|\theta, \omega]p_t(\theta, \omega)}{\sum_{(\theta', \omega')}p[o|\theta', \omega']p_t(\theta', \omega')}
\]

@@

Bayesian agents choose the effort level that maximizes their expected
flow payoff by taking expectations over their prior beliefs about
\(\theta\) and \(\omega\). Since agents are myopic, even though all the
parameters could be identified with enough variation in choices, there
will be instances in which the Bayesian will not learn even with
infinite amounts of data. This happens because, by being myopic, they do
not internalize the tradeoff between flow payoff and learning. This can
result in too little experimentation to learn their true type. An
alternative to this approach is given by \citet{Hestermann2021} and is
discussed in the concluding remarks. Regardless, this approach is useful
as a benchamrk for the other theories discussed in this paper.

@@

A Self-Serving Bayesian is an agent who uses a biased version of Bayes
rule to update their beliefs. They will update their beliefs about the
state \(\omega\) and his type \(\theta\) simultaneously by
over-attributing successes to a high value of \(\theta\) and
under-estimating the role of higher \(\omega\). Similarly, he will
attribute failure to a low state to a greater degree than an unbiased
agent would. To model the self-serving attribution bias, I take the
approach of \citet{benjamin2019}, where the posterior odds are given by:

@@

with
\(c(\theta_H, \omega, o)<c(\theta_M, \omega, o)<c(\theta_L, \omega, o)\leq 1\)
and
\(c(\theta, \omega_L, o)<c(\theta, \omega_M, o)<c(\theta, \omega_H, o)\leq 1\).

@@

@@

In order to compare the predictions of the theories discussed above, I
develop a unifying example that allows me to isolate the forces behind
each of the theories. The example is a modification of the one in
\citet{Heidhues2018} and is adapted to implementable in the lab.

@@

Denote the true type by \(\theta^*\) and the true state by \(\omega^*\).
The agent holds some prior belief about
\(\theta\)\footnote{which is potentially misspecified as in the dogmatic and switcher cases discussed above}
and chooses a binary gamble \(e in \{e_L, e_M, e_H\}\). The agent
observes whether the gamble is a success or a failure and gets a payoff
of \(1\) and if it is a success; they get \(0\) otherwise.

@@

I recruited 78 undergraduate subjects from the CESS lab at NYU who
participated in an in-person experiment. Sessions lasted approximately
45 minutes and subjects earned an average payment of \(\$22.6\). The
experiment was programmed using oTree \citep{otree}.

@@

In this section, I outline the behavior that is predicted by each of the
theories discussed above.

@@

@@

\textbf{Prediction 1S:} With positive probability, the switcher will
escape the self-defeating equilibrium and learn the true state.

One caveat is that when the switcher and the dogmatic agent both start
with a correctly specified belief, neither of them will fall into the
self-defeating equilibria and thus will look identical even in the long
run. This means that in order to distinguish between the two theories, I
need to look at cases where the agent starts with a misspecified belief.

@@

\textbf{Prediction 1A:} \emph{After observing a failure, the agent will
adjust their effort downwards by more than a Bayesian would have done.
After observing a success, the agent will adjust their effort upwards by
less than a Bayesian would have done. If the bias is large enough, the
agent might not adjust their effort upwards at all, or even decrease
their effort in response to a success.}

Prediction 1A helps distinguish between the self-attribution bias and
the dogmatic modeler. However, it does not help distinguish between the
self-attribution bias and the switcher. In order to do so, I need to
look at cases where the agent starts with a correctly specified belief.
In these cases, the switcher is very unlikely to switch to a different
belief and therefore, the behavior of the switcher will mostly change in
the same direction as the information he receives.
\footnote{If they get a streak of failures they are likely to 
adjust their effort downwards and if they get a streak of successes they are likely to adjust their effort upwards.}
Instead, the self-attribution bias will lead to a different behavior. In
particular, even for agents who have a belief that is correctly
specified and place a lot of weight on the correct values of \(\theta\),
the bias will cause these agents to become overconfident after a
sequence of successes. This force is not present in the switcher or the
dogmatic and it is the key feature that allows me to identify a biased
updating procedure.

@@

\hypertarget{the-bayesian-1}{%
\subsection{The Bayesian}\label{the-bayesian-1}}

The Bayesian serves as the closest benchmark to a fully rational agent.
However, as discussed above, they are still vulnerable to falling into
self-defeating equilibria, in which case, their long-run behavior will
be identical to that of the dogmatic modeler. The main distinguishing
feature is that when they do fall into a self-defeating equilibrium,
there is no bias in the direction of the misspecification that they end
up being trapped in. In contrast, the dogmatic modeler will always be
trapped in a misspecified belief that they started with and the
self-attribution updater will always become overconfident.

\textbf{Prediction 1B:} \emph{When an agent starts with a diffused prior
and falls into a self-defeating equilibrium, they will be equally likely
to fall into the underconfident trap as the overconfident trap.}

\textbf{Prediction 2B:} \emph{Bayesian agents always update their effort
choices in the same direction as the information they receive. That is,
if they see successes, they would only increase their effort; and if
they see failures they would only decrease their effort.}

@@

A closer look at whether people learned or not reveals that there is a
good amount of heterogeneity in the sample. Panel B of Figure
\ref{fig:learning-by-groups} shows the share of optimal choices by round
for subjects that chose an effort that matched the state in 3 out of the
last 4 rounds. It also shows the share of optimal choices for subjects
who chose an effort that matched the state in fewer than 3 out of the
last 4 rounds. I label the former as learners and the latter as
non-learners, with learners making up 39\% of the sample.

@@

Since the elicitation of the belief about the type was not incentivized
and not elicited in a standard way, I first need to confirm that
subjects were not just randomly choosing a matrix in which to enter
their effort. The left panel of Figure \ref{fig:matrix-choices-by-round}
shows the share of subjects who chose a matrix consistent with their
initial reported belief. In round 1, 71\% of the subjects chose a matrix
that was consistent with their initial belief. This indicates that
subjects were not just randomly choosing a matrix in which to enter
their effort. From round 2 onwards, the share steadily declines, but
still not as far as to indicate a random choice of matrices. This is
consistent with the subjects moving away from their starting belief
through some updating procedure.

@@

Overall, it does not seem to be the presence of traps that is preventing
learning. Panel A of Figure \ref{fig:traps-choice} shows that the share
of subjects who learned in cases where there was a trap is larger than
the share of subjects who learned in cases where there was no trap. This
is true for both treatments. This indicates that even in the presence of
traps, 44\% of subjects learn. From the share that did not learn when
there was a trap. A remaining 21\% can be accounted for as being
trapped.\footnote{A subject is considered trapped when 3 out of their last 4 choices are consistent with 
a self-defeating equilibrium.} While the rest of the subjects neither
learned nor were trapped.

@@

So far I have accounted for 16\% subjects as being trapped, and 39\%
having learned the state correctly. From the remaining 45\%, 60\% were
prone to traps but did not fall into them, therefore they could not have
been dogmatic modelers. Alternative explanations are that they were
switchers who considered an incorrect alternative model, or that they
were updating with a bias so large that their effort choices were far
from optimal. Behavior that is not consistent with either of these falls
outside of the scope of the theories discussed above.

@@

\begin{sidewaystable}[!htbp] \centering 
  \caption{Regression of effort change on news difference with robust standard errors} 
  \label{tab:regression} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{5}{c}{\textit{Dependent variable:}} \\ 
\cline{2-6} 
\\[-1.8ex] & \multicolumn{5}{c}{Change in effort} \\ 
 & All & Ego-relevant & Stereotype & Bayesian Simulation & Dogmatic Simulation \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4) & (5)\\ 
\hline \\[-1.8ex] 
 Good news & $-$0.12$^{**}$ & $-$0.16$^{***}$ & $-$0.05 & 0.08 & $-$0.08 \\ 
  & (0.05) & (0.05) & (0.05) & (0.05) & (0.05) \\ 
  & & & & & \\ 
 News difference & 0.22$^{***}$ & 0.22$^{***}$ & 0.21$^{***}$ & 0.06$^{***}$ & 0.10$^{***}$ \\ 
  & (0.02) & (0.02) & (0.02) & (0.02) & (0.02) \\ 
  & & & & & \\ 
 News difference * Good news & $-$0.27$^{***}$ & $-$0.25$^{***}$ & $-$0.29$^{***}$ & $-$0.04 & $-$0.06$^{***}$ \\ 
  & (0.02) & (0.02) & (0.02) & (0.02) & (0.02) \\ 
  & & & & & \\ 
 Constant & 0.31$^{***}$ & 0.31$^{***}$ & 0.30$^{***}$ & $-$0.08$^{*}$ & 0.05 \\ 
  & (0.04) & (0.04) & (0.04) & (0.04) & (0.04) \\ 
  & & & & & \\ 
\hline \\[-1.8ex] 
Observations & 4,680 & 2,700 & 1,980 & 4,680 & 4,680 \\ 
R$^{2}$ & 0.04 & 0.04 & 0.04 & 0.05 & 0.06 \\ 
Adjusted R$^{2}$ & 0.04 & 0.04 & 0.04 & 0.05 & 0.06 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{5}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{sidewaystable}

>>>>

\title{Learning with Mental Models: the case of overestimation}
\author{Jimena Galindo}
\date{October 24, 2023}

\begin{document}
\maketitle
\begin{abstract}
I design a framework and a laboratory experiment that allow for the
comparison of multiple theories of misspecified learning. I focus on a
framework with endogenous information and a data-generating process
ruled by two fundamentals: an ego-relevant parameter and a state. Within
this framework I study three forces that can lead to misspecified
beliefs: initial misspecifications, learning traps and biased updating.
I find that biased updating is the main driver of misspecified beliefs
in the lab. In addition, I vary the degree of ego-relevance of the
parameter by introducing a stereotype treatment. The data is consistent
with biased updating in both cases but for potentially different
reasons: when learning about themselves, subjects attribute successes to
their own ability and failures to luck. Instead, in the stereotype
treatment, they compensate for initial negative biases by
over-attributing positive signals to the ability of others. This
translates into similar observed choices but different dynamics in
beliefs.
\end{abstract}

@@

When the setting features learning traps, even an agent who incorporates
all information correctly, may fall into learning traps as outlined by
\citet{Hestermann2021}. These traps are characterized by a combination
of an incorrect belief and an optimal action which produce information
that confirms the incorrect belief. Once an agent falls into a trap, the
belief will be stable and even with a correctly specified model of
learning, they will not be able to abandon their misspecified beliefs.
If the agent is dogmatic about their initial belief,
\citet{Heidhues2018} show that they will inevitably fall into a trap and
thus will be able to rationalize and sustain their initial
misspecification
belief.\footnote{\citet{Gotte2022} study the case of agents with dogmatic initial beliefs in a 
laboratory experiment.}

\citet{Ba2023} moves away from dogmatism and endows the agents with a
mechanism through which they can abandon incorrect beliefs. This allows
them to avoid falling into learning traps. To do so agents perform
Bayesian hypothesis tests that evaluate which is the more likely
parameter out of two possibilities. By doing this, she characterizes the
set of situations in which, even agents who consider alternative
paradigms, may become trapped.\footnote{A 
similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion.}

@@

The agent chooses an action \(a\in A\) and observes a noisy outcome
\(h\). The outcome is a function of the agent's type, the state, and the
action. In particular \(h = h(\theta^*, \omega^*, a) + \varepsilon\)
with \(h(\cdot)\) increasing in both \(\theta^*\) and \(\omega^*\), and
such that conditional on a pair of parameters \((\theta, \omega)\),
there is a unique optimal action. Additionally,
\(\varepsilon\sim N(0, \sigma)\) is noise in the output.

After observing the outcome, the agent updates his beliefs about
\(\theta\) and \(\omega\) using some algorithm and moves on to the next
period. He repeats this process infinitely many times. I make the
simplifying assumption that the agent is myopic and chooses the action
that maximizes the payoff in each period. This assumption simplifies the
analysis and plays a role in whether an agent who updates their beliefs
using Bayes rule would learn the truth or not, however, for the main
theory discussed in this section, the results hold even when relaxing
this assumption.

A key notion in this setting is that of a self-defeating
equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in \citet{Esponda2016} to this setting with only one agent}.
A \emph{self-defeating equilibrium} is a belief and action pair,
\((p, a)\), such that the agent's belief about their type is
misspecified, and the outcome generated by the action is consistent with
the misspecified belief. The agent's belief is said to be \emph{stable}
when this happens and the average output is

\[\bar{h}(a^*) = h(\hat{\theta}, \hat{\omega}, a)\] Where
\(\hat{\theta}\) and \(\hat{\omega}\) is the type-state pair on which
the misspecified belief places all probability. That is, the agents
belief is \(p(\hat{\theta}, \hat{\omega}) = 1\) and
\(p(\theta, \omega) = 0\) for all other values of the parameter.

Within this framework, I consider two nested theories of belief
updating: the first one is a dogmatic modeler adapted from
\citet{Heidhues2018},The second one is a switcher, as in \citet{Ba2023}.
The dogmatic modeler can be seen as a switcher with an infinitely sticky
initial belief so the two theories are nested. However, outside of the
limiting case, they produce different predictions about the agent's
behavior. I discuss each of them separately in what follows.

@@

Take the action space to be the same as the state space \(\Omega\) and
let \(\Omega=A:=[\underline{a}, \infty)\).

A dogmatic agent is characterized by a learning model where he does not
update their beliefs about \(\theta\), instead, he holds a degenerate
belief that places all probability on \(\hat{\theta}\), which is
potentially misspecified. In this case, no matter how much information
he gathers against being of type \(\hat{\theta}\), he will not update
his beliefs about it. Any discrepancies between the observed outcomes
and his believed type are incorporated using the Bayes rule to update
his beliefs about \(\omega\). \citet{Heidhues2018} show that, under
certain assumptions on the per-period
utility,\footnote{The assumptions are that $u$ is twice continuously differentiable with: (i)$u_{aa}<0$ and $u_a(\underline{a},  \theta, \omega)>0>u_a(\Bar{a}, \theta, \omega)$, (ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{a\theta}<0$ and $u_{a\omega}>0$. Where $\underline{a}$ is the
minimal action and $\Bar{a}$ is the maximal one. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} a dogmatic
modeler will inevitably fall into a self-defeating equilibrium. The
equilibrium will be such that the outcomes they observe reinforce their
belief on \(\omega\) in such a way that as \(t\to\infty\) the agent will
be sure that the state is some \(\omega^{'}\) consistent with their
believed type and the observed data. In other words, they will be in a
self-defeating equilibrium with a stable belief that places probability
one on the incorrect parameters \((\hat{\theta}, \omega_{\infty})\)

The mechanism by which the dogmatic agent falls into the self-defeating
equilibrium is the following: Suppose the agent holds the misspecified
belief that they are type \(\hat{\theta}>\theta^*\). For any prior over
\(\omega\), the agent will be disappointed by the outcome. That is:
\[\bar{h}(\theta^*, \omega^*, a_t^*) := \frac{1}{T}\sum_t \left[h(\theta^*, \omega^*, a_t^*)+\varepsilon_t\right]<\frac{1}{T}\sum_t \left[h(\hat{\theta}, \mathbb{E}(\omega), a_t^*)+\varepsilon_t\right]\]
Where the right-hand-side of the inequality represents the average
outcome that the agent expected.

There are two possible sources for the disappointment: the first is that
the realized state is lower than the expected state; the second source
is that the agent is of type \(\theta^*\) and therefore, for all
possible states, his gain will be lower than what he expected.

Because the agent is dogmatic, he will not update his beliefs about
\(\theta\) and as a consequence will attribute the disappointment to the
state being lower than expected. He will continue to update in this way
until he converges to a belief about \(\omega\) that is stable. Such a
belief will explain the observed utility perfectly and allow the agent
to rationalize his dogmatic belief about \(\theta\). Under the
assumptions of \citet{Heidhues2018}, there is a unique value of
\(\omega\) at which the belief is stable, I will refer to such value as
\(\omega_\infty\). This mechanism is illustrated in Example 1.

@@

Where \(h^t\) is the history of outcomes up to time \(t\), and
\(\alpha\) is the switching
threshold.\footnote{Notice that if $\alpha \to \infty$, the behavior of the switcher will be indistinguishable from that of the Dogmatic modeler. 
In this sense, the switcher is a generalization of the dogmatic type.}
By keeping track of the likelihood ratio, the agent can perform a
\emph{Bayesian hypothesis test} and adopt the Dogmatic belief that best
fits the
data.\footnote{In a related problem \citet{Schwarstein2021} proposes a similar updating procedure which relies on the Bayesian hypothesis test. However, in their model there is a sender who optimally chooses to propose a model that fits the data}

@@

A Bayesian agent simultaneously updates their beliefs about \(\theta\)
and \(\omega\) by using Bayes' rule. The posterior at period t+1 about
\(\theta\) after observing outcome \(o\) is given by: \[
p_{t+1}(\theta, \omega| h^t) = \frac{p[o|\theta, \omega]p_t(\theta, \omega)}{\sum_{(\theta', \omega')}p[o|\theta', \omega']p_t(\theta', \omega')}
\]

@@

Bayesian agents choose the effort level that maximizes their expected
flow payoff by taking expectations over their prior beliefs about
\(\theta\) and \(\omega\). Since agents are myopic, even though all the
parameters could be identified with enough variation in choices, there
will be instances in which the Bayesian will not learn even with
infinite amounts of data. This happens because, by being myopic, they do
not internalize the trade off between flow payoff and learning. This can
result in too little experimentation to learn their true type. An
alternative to this approach is given by \citet{Hestermann2021} and is
discussed in the concluding remarks. Regardless, this approach is useful
as a benchmark for the other theories discussed in this paper.

@@

A Self-Serving Bayesian is an agent who uses a biased version of Bayes
rule to update their beliefs. They will update their beliefs about the
state \(\omega\) and his type \(\theta\) simultaneously by
over-attributing successes to a high value of \(\theta\) and
under-estimating the role of higher \(\omega\). Similarly, he will
attribute failure to a low state to a greater degree than an unbiased
agent would. This sort of biased updating is referred to as
\emph{misattribution bias}.

To model the misattribution bias, I take the approach of
\citet{benjamin2019}, where the posterior odds are given by:

@@

with
\(c(\theta_H, \omega, o)<c(\theta_M, \omega, o)<c(\theta_L, \omega, o)\)
and
\(c(\theta, \omega_L, o)<c(\theta, \omega_M, o)<c(\theta, \omega_H, o)\).
And where \(h_t\) is the history of signal realizations observed up to
period t.

@@

Define the distortion in the likelihood by
\(\Delta_c p(h^t|\theta, \omega) = p(h^t|\theta, \omega)^{c(\theta, \omega, o_t)} - p(h^t|\theta, \omega)\).
Notice that because \(p(h^t|\theta, \omega)<1\), if \(c(\cdot)<1\), a
smaller exponent \(c\) implies a larger distortion upwards for fixed
\(h^t\), and fixed parameters \(\theta\) and \(omega\). While a
\(c(\cdot)>1\) implies a distortion downwards and it is larger when
\(c(\cdot)\) is larger.

@@

In order to compare the predictions of the theories discussed above, I
develop a unifying example where the forces behind each of the theories
can be isolated. The example is a modified from \citet{Heidhues2018} and
is adapted to be implementable in the lab.

@@

Denote the true type by \(\theta^*\) and the true state by \(\omega^*\).
The agent holds some prior belief about
\(\theta\)\footnote{which is potentially misspecified as in the dogmatic and switcher cases discussed above}
and chooses gamble \(e in \{e_L, e_M, e_H\}\). The agent observes
whether the gamble is a success or a failure and gets a payoff of \(1\)
and if it is a success; they get \(0\) otherwise.

@@

I recruited 86 undergraduate subjects from the CESS lab at NYU who
participated in an in-person experiment. Sessions lasted approximately
45 minutes and subjects earned an average payment of \(\$22.6\). The
experiment was programmed using oTree \citep{otree}.

@@

In this section, I outline the behavior that is predicted by each of the
theories discussed above. Some of the predictions are made relative to
the Bayesian benchmark. In particular, the reaction to good and bad news
is very informative of which model is most likely to be underlying the
behavior. Define \emph{good news} to be a sample of signal realizations
such that the number of successes in the sample is larger than the
average number of successes observed so far. i.e.~if

\[ \sum_i \mathbb{I}\{o_{iT}=success\} \geq \frac{1}{T} \sum_{t=0}^T\sum_i \mathbb{I}\{o_{it}=success\}\]
where T is the current period and \(o_{t}=[o_{it}]_{i=1}^I\) is the
sample observed at period \(t\). \emph{Bad news} are any other kind of
signal realization. Denote the set of good news at period t by
\(\mathcal{G}_t\) and let \(s_t\) be the number of successes in \(o_t\).

Given the definition of good and bad news, the reaction of the agent to
each kind of news is the change in their choice, conditional on the
signal realization. Therefore, what we are interested are the
coefficients of the following regression model, where the actions are
those that a bayesian agent would choose:

\[ \Delta a = \beta_0 + \beta_1\mathbb{I}\{o_t\in \mathcal{G}_t\} + \beta_2(s_{t} - s_{t-1}) + \beta_3\mathbb{I}\{o_t\in \mathcal{G}_t\}(s_{t} - s_{t-1})  \]

Where \((s_{t} - s_{t-1})\) is the size of the surprise and \(\Delta a\)
is the change in action. \(\beta_2\) is the reaction of the agent to bad
news with surprise \((s_{t} - s_{t-1})\). And \(\beta_2 + \beta_3\) is
the reaction to good news with surprise \((s_{t} - s_{t-1})\). Notice
that for a dogmatic modeler \(\beta_2\geq0\) and
\(\beta_2+\beta_3\geq 0\) as more successes are always indicative of a
better state.

@@

In term of the dogmatic's reagtion to good and bad news, let \(\beta^D\)
denote the coefficients of the regression model above when the actions
are those predicted by the dogmatic model.

\textbf{Prediction 2D:} \emph{Before falling into a trap, the dogmatic
modeler reacts more to good and bad news than the Bayesian would.
i.e.~\(\beta_2^D>\beta_2\) and \(\beta_3^D > \beta_3\).}

@@

\textbf{Prediction 1S:} \emph{With positive probability, the switcher
will escape the self-defeating equilibrium and learn the true state.}

Let \(\beta^S\) be the coefficients of the regresion model when the
actions are those taken by a switcher.

\textbf{Prediction 2S:} \emph{When there is no change of paradigm, the
switcher behaves as the dogmatic. When there is a change in paradigm,
the reaction to good and bad news is attenuated or even in the opposite
direction of what a Bayesian would have done. that is
\(\beta_2^S<\beta_2\) and \(\beta_2^S + \beta_3^S<\beta_2+\beta_3\)}

One caveat is that when the switcher and the dogmatic agent both start
with a correctly specified belief, neither of them will fall into the
self-defeating equilibria and thus will look identical even in the long
run. This means that in order to distinguish between the two theories, I
need to look at cases where the agent starts with a misspecified belief
and the switcher changes paradigm.

@@

Let \(\beta^A\) be the coefficients of the regression model when the
agent updates with misattribution bias.

\textbf{Prediction 1A:} \emph{After observing a failure, the agent will
adjust their effort downwards by more than a Bayesian would have done.
After observing a success, the agent will adjust their effort upwards by
less than a Bayesian would have done. If the bias is large enough, the
agent might not adjust their effort upwards at all, or even decrease
their effort in response to a success. That is \(\beta_2^A > \beta_2\)
and \(\beta_2^A + \beta_3^A<\beta_2 + \beta_3\)}

In addition to their behavior, there is an additional characteristic
feature of misattribution bias that is not present in other models. This
is the only model in which an agent could start off with a correct
belief and become more overconfident. In the long-run.

@@

@@

A closer look at whether people learned or not reveals that there is a
good amount of heterogeneity in the sample. Panel B of Figure
\ref{fig:learning-by-groups} shows the share of optimal choices by round
for subjects that chose an effort that matched the state in 3 out of the
last 4 rounds. It also shows the share of optimal choices for subjects
who chose an effort that matched the state in fewer than 3 out of the
last 4 rounds. I label the former as learners and the latter as
non-learners, with learners making up 38\% of the sample.

@@

Since the elicitation of the belief about the type was not incentivized
and not elicited in a standard way, I first need to confirm that
subjects were not just randomly choosing a matrix in which to enter
their effort. The left panel of Figure \ref{fig:matrix-choices-by-round}
shows the share of subjects who chose a matrix consistent with their
initial reported belief. In round 1, 69\% of the subjects chose a matrix
that was consistent with their initial belief. This indicates that
subjects were not just randomly choosing a matrix in which to enter
their effort. From round 2 onwards, the share steadily declines, but
still not as far as to indicate a random choice of matrices. This is
consistent with the subjects moving away from their starting belief
through some updating procedure.

@@

Overall, it does not seem to be the presence of traps that is preventing
learning. Panel A of Figure \ref{fig:traps-choice} shows that the share
of subjects who learned in cases where there was a trap is larger than
the share of subjects who learned in cases where there was no trap. This
is true for both treatments. This indicates that even in the presence of
traps, 43\% of subjects learn. From the share that did not learn when
there was a trap. A remaining 22\% can be accounted for as being
trapped.\footnote{A subject is considered trapped when 3 out of their last 4 choices are consistent with 
a self-defeating equilibrium.} While the rest of the subjects neither
learned nor were trapped.

@@

So far I have accounted for 17\% subjects as being trapped, and 38\%
having learned the state correctly. From the remaining 45\%, 60\% were
prone to traps but did not fall into them, therefore they could not have
been dogmatic modelers. Alternative explanations are that they were
switchers who considered an incorrect alternative model, or that they
were updating with a bias so large that their effort choices were far
from optimal. Behavior that is not consistent with either of these falls
outside of the scope of the theories discussed above.

@@

\begin{sidewaystable}[!htbp] \centering 
  \caption{Regression of effort change on news difference with robust standard errors} 
  \label{tab:regression} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{5}{c}{\textit{Dependent variable:}} \\ 
\cline{2-6} 
\\[-1.8ex] & \multicolumn{5}{c}{Change in effort} \\ 
 & All & Ego-relevant & Stereotype & Bayesian Simulation & Dogmatic Simulation \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4) & (5)\\ 
\hline \\[-1.8ex] 
 Good news & $-$0.11$^{**}$ & $-$0.16$^{***}$ & $-$0.04 & 0.09$^{*}$ & $-$0.08 \\ 
  & (0.05) & (0.05) & (0.05) & (0.05) & (0.05) \\ 
  & & & & & \\ 
 News difference & 0.22$^{***}$ & 0.22$^{***}$ & 0.21$^{***}$ & 0.06$^{***}$ & 0.10$^{***}$ \\ 
  & (0.02) & (0.02) & (0.02) & (0.02) & (0.02) \\ 
  & & & & & \\ 
 News difference * Good news & $-$0.27$^{***}$ & $-$0.25$^{***}$ & $-$0.29$^{***}$ & $-$0.03 & $-$0.06$^{***}$ \\ 
  & (0.02) & (0.02) & (0.02) & (0.02) & (0.02) \\ 
  & & & & & \\ 
 Constant & 0.31$^{***}$ & 0.31$^{***}$ & 0.30$^{***}$ & $-$0.09$^{**}$ & 0.06 \\ 
  & (0.04) & (0.04) & (0.04) & (0.04) & (0.04) \\ 
  & & & & & \\ 
\hline \\[-1.8ex] 
Observations & 5,160 & 2,700 & 2,460 & 5,160 & 5,160 \\ 
R$^{2}$ & 0.04 & 0.04 & 0.04 & 0.05 & 0.06 \\ 
Adjusted R$^{2}$ & 0.04 & 0.04 & 0.04 & 0.05 & 0.06 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{5}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{sidewaystable}
